---
format: 
    revealjs:
        css: "include/style.css"
        chalkboard: true
        auto-stretch: false
        multiplex: true
        slide-number: c/t
        mermaid:
          theme: neutral
        width: 100%
        code-fold: true
        code-tools: true
        code-overflow: wrap
bibliography: references.bib
title: "Interactions and Moderation"
subtitle: DClin Research Methods 1
author: Dr Christopher Wilson
institute: Teesside University
logo: "logo.jpg"
csl: apa.csl
---

## Recap {.smaller}

::: {.fragment .fade-in-then-semi-out .smaller}
- Thinking about more than outcomes. Designing studies to answer more **specific research questions / think about process**.
  - "Why is this happening?"
  - "What is the mechanism?"
:::

::: {.fragment .fade-in-then-semi-out}
- Thinking beyond significance testing. Using **confidence intervals and effect sizes** to interpret results.
  - "How big is the effect?"
  - "What is the range of plausible values?"
:::

::: {.fragment .fade-in-then-semi-out}
- Thinking about the relationship between variables. Modelling relationships between variables using regression.
  - "Does **Predictor Variable** (e.g. Treatment Group, Avoidance, Trait) predict **Outcome Variable** (e.g. Wellbeing, Depression, Behaviour)?"
  - "How much variance is explained by the model?"
:::


## Last week and the week before

::: {.callout icon=false}
- Thinking about the relationship between variables. Modelling relationships between variables using regression.
  - "Does **Predictor Variable** (e.g. Treatment Group, Avoidance, Trait) predict **Outcome Variable** (e.g. Wellbeing, Depression, Behaviour)?"
  - "How much variance is explained by the model?"
:::

- We looked at building regression models based on our research questions and hypotheses.

- This allowed us to consider the contribution of multiple predictors to an outcome variable.


## Today

::: {.callout icon=false}
- Thinking about more than outcomes. Designing studies to answer more **specific research questions / think about process**.
  - "Why is this happening?"
  - "What is the mechanism?"
:::

>- We will be looking at how we can use regression to model **interactions** between variables.

>- We will also be looking at how we can use regression to model **moderation** of relationships between variables.


## What are interactions? #1

- Interactions are when the outcome variable is influenced by the interaction between two or more predictor variables.

For example, if we are looking at whether *Time in counselling* and *Rapport level with clinician* predict *General Wellbeing*.

- Without the interaction, we consider the effect of *Time in counselling* or *Rapport level with clinician* on *General Wellbeing*.

- With the interaction, we consider the effect of *Time in counselling* on *General Wellbeing* at different levels of *Rapport level with clinician* (or vice versa).



```{r echo=F, eval = T}

# load datasets/modData.csv

modData <- read.csv("../../datasets/modData.csv")


```

# Note: All assumptions of multiple regression also apply to interactions and moderation.


## How do we check for interactions? {.smaller}

- We can check if there is an interaction between predictor variables when we build our regression model.

- We do this by adding a term to our model that represents the interaction between the predictor variables.

```{r eval = F, echo = T}

# model1 without interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling + rapportLevel) #<1>

# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel) #<2>

```

1. This model does not include an interaction between timeInCounselling and rapportLevel. This is because we used the + symbol to separate the predictor variables.

2. This model does include an interaction between timeInCounselling and rapportLevel. This is because we used the * symbol to separate the predictor variables.

## Interpreting interactions in regression #1 {.smaller}

```{r eval = T, echo = T}

# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel) #<2>

summary(model1)

```

## Interpreting interactions in regression #2 

- We can see that the overall model is significant

- We can see that the interaction between timeInCounselling and rapportLevel is significant

- We can see that the main effects of timeInCounselling and rapportLevel are not significant

- In models where the interaction is significant, we should not interpret the main effects of the predictor variables (even when they are significant).

## Interpreting interactions in regression #3 

- We can look at the **simple slopes** of the interaction.

```{r eval = T, echo = F}

library(sjPlot)

plot_model(model1, type="int", mdrt.values = "meansd")

#We are looking at different levels of rapportLevel. We are specifying that we want to plot the interaction at the mean and +/- 1SD of rapportLevel (`mdrt.values = "meansd"`).
```

## Interpreting interactions in regression #4

- We can further test this interaction by calculating range of values where the interaction is significant.

- We can do this using the `rockchalk` package. We will do it our moderation example later.


## Thinking about interactions

- We can see that the relationship between timeInCounselling and generalWellbeing is different at different levels of rapportLevel.

- This means that the variables are interacting with each other to predict the outcome variable.

::: {.callout-warning}
If we plan to test for interactions, we need to make sure that we have enough power to do so. It is like adding another predictor variable to the model.
:::

# Moderation designs 

## What are moderation designs? 

- Moderation designs are when the relationship between two variables is moderated by a third variable.

- This is similar to an interaction, but we are stating our theoretical model about the relationship between the variables in a more specific way.

- For example, we might be interested in whether the relationship between Trait Anxiety and Depression is moderated by Avoidance.

- We are specifically stating that we believe Trait Anxiety predicts Depression, but that their relationship is different at different levels of Avoidance.

# So moderation is just an interaction effect?

Statistically, yes. Theoretically, no.

From a theory perspective, we are stating that we believe that the relationship between two variables is moderated by a third variable.

## Diagram of moderation design

```{r fig.height = 5, fig.width = 10, fig.align='center', echo=F}


#| fig.cap="Diagram of moderation design"
#| fig.asp=0.5
#| fig.pos="H"
#| fig.align='center'

library(DiagrammeR)

code <- "
	digraph moderation_model { 
	  # Nodes  
	  node [shape = rectangle, fontname = Arial, fontsize=9]
	  iv    [pos = '-1,1!', label = 'Time in\ncounselling']
	  blank [pos =  '0,1!', label = '', width = 0.001, height = 0.001]
	  mod   [pos =  '0,2!', label = 'Rapport\nlevel']
	  dv    [pos =  '1,1!', label = 'General\nwellbeing']
	    
	  # Hide moderator edges
	  { rank = same; blank mod }
	  
	  # Edges
	  edge [fontname = Arial, fontsize=8]
	  mod -> blank
	  iv -> blank [dir = none]
	  blank -> dv [minlen = 1]

	  # Graph
	  graph [layout = neato, overlap = true, outputorder = edgesfirst]
}"

grViz(code)

```

## When should we use moderation designs?

- We should use moderation designs when we have a theoretical reason to believe that the relationship between two variables is moderated by a third variable.

- This should come from a good understanding of the literature and the theory behind the variables.

- We should not use moderation designs "just to see what happens". The statistical test alone does not **prove** the existence of the theoretical relationship.


## How do we check for moderation? 

- The first step is to build a regression model that includes the main effects and the interaction between the two predictor variables (this is the same as interaction that we covered earlier).

- If the interaction is significant, we can then visualise the interaction using simple slopes (same as what we did earlier).

- We can then calculate the range of values of the moderator variable where the interaction is significant.

## Moderation example

- Let's imagine that we theorise that the relationship between *Time in counselling* and *General Wellbeing* is moderated by *Rapport level*.

- We want to know at which levels of *Rapport level* the relationship between *Time in counselling* and *General Wellbeing* is significant.

- We also want to know the direction of the relationship between *Time in counselling* and *General Wellbeing* at different levels of *Rapport level*.

## Moderation example #2

The regression model is the same as the one we used for interactions.

```{r eval = F, echo = T}

# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel) #<1>

```

1. This model includes an interaction between timeInCounselling and rapportLevel. We know this because we used the * symbol to separate the predictor variables.


## Testing the moderation effect 

Testing at which levels of *Rapport level* the relationship between *Time in counselling* and *General Wellbeing* is significant.

Step 1: Run the regression model:


```{r eval = T, echo = T}


# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel)


```

## Testing the moderation effect #2 

Step 2: Plot the interaction:


```{r eval = T, echo = T}



# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel)

# using the rockchalk package

library(rockchalk)

# use the plotSlopes() function

ps <- plotSlopes(model1, plotx = "timeInCounselling", modx = "rapportLevel", interval = "confidence", modxVals = "std.dev")

ps

```

## Testing the moderation effect #3 {.scrollable} 

Step 3: Test the range of values where the interaction is significant: 


```{r eval = T, echo = T}
 

# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel)

# using the rockchalk package

library(rockchalk)

# use the plotSlopes() function

ps <- plotSlopes(model1, plotx = "timeInCounselling", modx = "rapportLevel", interval = "confidence", modxVals = "std.dev") 

# use the testSlopes() function

ts <- testSlopes(ps)



```

## Testing the moderation effect #4 {.scrollable}

Step 4: Visualise the results:


```{r eval = T, echo = T}
 

# model1 with interactions

model1 <- lm(data = modData, generalWellbeing ~ timeInCounselling * rapportLevel)

# using the rockchalk package

library(rockchalk)

# use the plotSlopes() function

ps <- plotSlopes(model1, plotx = "timeInCounselling", modx = "rapportLevel", interval = "confidence", modxVals = "std.dev")

# use the testSlopes() function

ts <- testSlopes(ps)

# plot the range of significant values

plot(ts)

```

## How do we interpret the moderation effect?

- We can see that the relationship between *Time in counselling* and *General Wellbeing* is significant levels of *Rapport level* outside of -10.8 and 3.54.

- However, the *Rapport level* scale does not actually go below 0. So we can say that the relationship between *Time in counselling* and *General Wellbeing* is significant when *Rapport level* is above 3.54.

- We can look at the first interaction plot to see the direction of the relationship between *Time in counselling* and *General Wellbeing* at increasing levels of *Rapport level*.

## Summary

- We can use regression to model interactions between predictor variables.

- We can also use regression to model moderation of relationships between predictor variables.

- Moderations are just a specific type of interaction. They are based on specific theoretical models about the relationship between variables.

- We can use the `rockchalk` package to test for interactions and moderation and plot them.




