
---
title: Practical 7 Factor Analysis
author: "Dr. Christopher Wilson"
Subtitle: "PSY4179 Research Methods 1"
Institution: "Teesside University"
format:
  html:
    css: ["../include/style.css","../include/glossary.css","../include/booktem.css", "../include/webex.css"]
    include-in-header: ["../include/webex.js"]
execute:
    eval: false
    echo: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(psych)
library(tidyverse)
library(GPArotation)
```

# Introduction

Before we begin, install and load the following packages:

- tidyverse
- psych
- GPArotation

For this session, we will use the BFI dataset from the psych package. It is made up of 25 self-report personality items from the International Personality Item Pool, gender, education level and age for 2800 subjects and used in the Synthetic Aperture Personality Assessment.

The personality items are split into 5 categories: Agreeableness (A), Conscientiousness (C), Extraversion(E), Neuroticism(N), Openness(O). 

Each item was answered on a six point scale: 1 Very Inaccurate, 2 Moderately Inaccurate, 3 Slightly Inaccurate, 4 Slightly Accurate, 5 Moderately Accurate, 6 Very Accurate.

::: {.callout-note}
# Task 1: Load the data

To load the data file, use the following code:

```{r}
#| echo: true
#| eval: false

 data("bfi")

```

The data will be loaded into the environment tab as a temporary file. You need to click on it to make use of it in your code.

:::

::: {.callout-note}

# Task 2: Filter the data

Columns 1 to 25 are the data we are interested in. To include only these columns in analysis, we use the code below to make a subset of the data 



```{r}
#| echo: true
#| eval: false

bfidata <- bfi[1:25] %>% na.omit() #<1>

```
1. The code above creates a new data frame called bfidata that contains only the columns we are interested in (1:25) and removes any rows with missing data (na.omit())

:::

::: {.callout-note}

# Task 3: Examine the data

1. Use the describe() command to review a summary of the data


```{r}
#| echo: true
#| eval: false

describe(bfidata)
```



2. Check the number of complete responses using the commands: 
```{r}
#| echo: true
#| eval: false

bfidata %>% complete.cases() %>% sum()
```

:::


::: {.callout-note}

# Task 4: Check the assumptions of factor analysis

- Run Bartletts test of sphericity

```{r}
# make a correlation matrix first

bfi.maxtrix <- cor(bfidata)

# run bartlettt's test
cortest.bartlett(bfi.maxtrix, n= 2436)

```


- Run the Kaiser-Meyer-Olkin (KMO) measure of sampling adequacy

```{r}
KMO(bfidata) # using the data

#OR 

KMO(bfi.maxtrix) # using the correlation matrix
```

:::

::: {.callout-note}

# Task 5: Run a factor analysis


- Run a parallel analysis to determine the number of factors

```{r}
fa.parallel(bfidata)
```


-  Run a factor analysis based on the suggested number of factors (no rotation)

```{r}
factoranalysis1 <- fa(bfidata, nfactors = 6, fm="pa", max.iter = 100, rotate = "none")
```


- Interpret the output and determine an optimal number of factors based on: scree plot, variance levels, eigenvalues

- Re-run factor analysis with optimal number of factors and use "oblimin" rotation

:::

```{r}
factoranalysis2 <- fa(bfidata, nfactors = 5, fm="pa", max.iter = 100, rotate = "oblimin")

fa.diagram(factoranalysis2) ## can use fa.diagram to see factors and the questions that load onto them

```


# You are done for today {{< fa hands-clapping >}}

::: {.callout-tip collapse="true"}
# Further resources for  factor analysis

- [Factor Analysis : Andy Field](https://study.sagepub.com/sites/default/files/labcoatleni_worldwide_addiction.pdf) based on SPSS but useful for understanding the process

- [Factor Analysis : Andy Field](https://www.discovr.rocks/solutions/code/code_18/) A code example with no explanation but useful for checking your code

- [Factor analysis](https://statisticsbyjim.com/basics/factor-analysis/) A useful basic explanation of factor analysis



:::
