---
title: Practical 6 Mediation Analysis
author: "Dr. Christopher Wilson"
Subtitle: "PSY4179 Research Methods 1"
Institution: "Teesside University"
format:
  html:
    css: ["../include/style.css","../include/glossary.css","../include/booktem.css", "../include/webex.css"]
    include-in-header: ["../include/webex.js"]
---


```{r setup, include=FALSE}
library(webexercises)
```


::: {.callout-tip collapse="true"} 
# How to approach this practical session

1. Have RStudio open and ready to use.
1. Be sure to read the contextual information and instructions carefully. This information is necessary to complete the tasks.
1. Complete the tasks as you go along - they are labelled "Task".
1. Some of the code you need will be completed, some will be partially completed, some you will need to do yourself. You can copy code examples from this page by clicking on the clipboard icon {{< fa clipboard >}} which appears when you hover over the code examples.
1. Buttons to press in your keyboard are shown like this: <pc>Ctrl + S</pc> or <mac>Cmd + S</mac>. 
1. Menu items to click in R Studio are shown like this: <if>File -> Save</if>. Panes or Tabs in R Studio are also shown like this: <if>Environment</if>.
1. On this page, boxes with a drop down arrow {{< fa square-caret-down >}} can be collapsed to hide the information inside them. You can also click on the {{< fa square-caret-right >}} arrow to expand the section.
:::


::: {.callout-important collapse="false"}

# Download the data for this practical

For this session you will need to download the following dataset: [practical6_data.csv](practical6_data.csv)

Right click and select <if>Save link as...</if> to download the file. Save it in the same folder as your script.

### Reminder: Importing data into R

To import data into R, you can use the a command that relates to the type of data you are importing. For example, to import the data in the file **practical6_data.csv**, you would use the following command:

```{r eval=FALSE}
practical6_data <- read.csv("practical6_data.csv")

```

You can also use the menu to import data. To do this, click on the <if>Import Dataset</if> button and select the type of data you want to import. For example, to import the data in the file **practical6_data.csv**, you would click on <if>From Text (readr)</if> and then select the file **practical6_data.csv**.

:::

## Import the data

::: {.callout-note}

# Task

Import the data from the file **practical6_data.csv** into R. Save the data in a variable called **practical6_data**.

Check the data has been imported correctly
:::

## Overview of research 
In these exercises, we will work with the dataset from Tal-Or et al. (2010). Tal-Or et al. (2010) examined the presumed effect of the media in two experimental studies. These data are from study 2.

### Abstract
According to the influence of presumed media influence hypothesis, people estimate the potential effects of media on other people and change their attitudes or behaviours as a consequence. In recent years, many studies offered some support for this idea. However, a
central limitation of these studies is that all of them utilized correlational methodology and thus do not offer a valid way to infer causality. The current research examined the causal direction in the influence of presumed media influence using experimental methodology. In Study 2,perceptions regarding the influence of a news story about an expected shortage in sugar were manipulated indirectly, by manipulating the perceived exposure to the news
story,and behavioural intentions resulting from the story were consequently measured.

### Experimental hypothesis 

Participants will be more likely to react to the sugar shortage report when they are manipulated to think that others are more affected by the report.

### Procedure
Experimenters explained that participants were to answer a short survey about media impact on purchase decisions and handed out randomly two versions of a two page pamphlet. In the first page, participants were asked to read a newspaper article about an expected shortage in sugar that is about to be published in the coming days. In the second page, the participants were asked to answer three questionnaires, one measuring reactions to the publication of the shortage in sugar story, the second served as a manipulation check, and the final set of questions were demographic. After respondents completed the questionnaires, they were debriefed as to the true object of the study and about the fact that the information they received about shortage in sugar was false.

### Data

The results are data frame with 123 observations on the following 6 variables:

- **cond**: Experimental Condition: 0 low media importance, 1 high media importance. (Importance was manipulated by telling participants that the story would be published either on the front page of a newspaper or inside the economic supplement of the newspaper).

- **pmi**: Presumed media influence (This was a manipulation check to see that participants felt that publishing the article would affect people’s behaviour and that responses were not different based on age or gender of participants. The value is the mean score of response to two questions.) 

- **import**: Importance of the issue (Similar to the **pmi** variables, a check was also done to see that participants agreed that the issue is important and that responses were not different based on age or gender of participants. “The article is about an important topic concerning the global economic crisis” and “the decision to publish the article on the front/back page is an indication of the topic’s importance”.)

- **reaction**: Subjects rated agreement about possible reactions to the story. (The score is the mean of 4 items: “publishing the article might affect my decisions about buying sugar, “I will buy sugar next time I go to the store,” “I will buy more sugar than usual next time I go to the store,” and “I will try to reduce my sugar consumption.”).

- **gender**: 1 = male, 2 = female

- **age**: a numeric vector

::: {.callout-note}
# Task:
Run a mediation analysis to see whether the relationship between **cond** and **reaction** is mediated by **pmi**.

1.  Use Baron and Kenny (1986) Approach
2.  Use Preacher and Hayes (2008) Approach

- For each analysis, identify if there is a mediation effect and explain why.

You will need to use the following packages (install them first, if necessary):
```r
    library(mediation) #Mediation package
    
    library(multilevel) #Sobel Test
    
    library(bda) #Another Sobel Test option
    
    library(gvlma) #Testing Model Assumptions 
    
    library(stargazer) #Handy regression tables
```
:::


# You are done for today {{< fa hands-clapping >}}

::: {.callout-tip collapse="true"}
# Further resources for mediation analysis

[DClin Research Methods 1 - Mediation Analysis](https://christopherjwilson.github.io/APRM23/mediation.html)

[R: a language not a letter: mediation](https://ademos.people.uic.edu/Chapter14.html)

[Testing indirect effects in R](https://nmmichalak.github.io/nicholas_michalak/blog_entries/2018/nrg01/nrg01.html)
:::
